---
title: "openalexVectorComp Demo Analysis"
subtitle: "Reference vs Corpus Workflow"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    df-print: paged
    code-fold: show
    code-summary: "Show code"
    anchor-sections: true
    embed-resources: true
    theme: flatly
    include-in-header:
      text: |
        <style>
        main.content {
          max-width: 1100px;
        }
        .quarto-title-block .quarto-title-banner {
          background: linear-gradient(120deg, #16335b 0%, #1f4f8a 100%);
          color: #ffffff;
        }
        .callout {
          border-radius: 8px;
        }
        .cell-output pre, pre {
          white-space: pre-wrap;
          overflow-wrap: anywhere;
          word-break: break-word;
        }
        .cell-output pre {
          background: #eef4ff;
          border: 1px solid #d4e2ff;
          border-radius: 6px;
          padding: 0.75rem;
        }
        </style>
execute:
  echo: true
  warning: true
  message: true
---

This demo runs a complete reference-vs-corpus workflow from the current
project directory. It is designed to be inspectable, so each step writes files
to disk and keeps them for review.

```{r demo-version-info, echo=FALSE, eval=TRUE}
pkg_version <- tryCatch(
  as.character(utils::packageVersion("openalexVectorComp")),
  error = function(e) "development"
)
cat("**openalexVectorComp version:** ", pkg_version, "\n\n", sep = "")
```

# 1. Setup

In setup we load the package, read backend configuration, and verify that both
`corpus/` and `reference_corpus/` are present. This ensures later steps fail
early with clear causes if inputs are missing.

::: {.callout-note}
This section checks inputs only. No embeddings are created yet.
:::

```{r setup-environment}
library(openalexVectorComp)

backend <- embedding_backend_read(file.path(".", "demo_backend.yaml"))

project_dir <- file.path(".", "project")
corpus_dir <- file.path(project_dir, "corpus")
reference_dir <- file.path(project_dir, "reference_corpus")

stopifnot(dir.exists(corpus_dir))
stopifnot(dir.exists(reference_dir))
```

## Model Configuration

This demo reads backend configuration from `demo_backend.yaml` in the demo
workspace (`demo_dir`). The file is created by
`run_demo_openalex_quarto()` using `embedding_backend_save()`.

After embeddings are generated, the effective model metadata used for the run
is written to:

- `project/embeddings/model_id=<...>/embed_model.yaml`

### Current backend in this demo

```{r model-configuration-info}
info <- embedding_backend_info(backend)

data.frame(
  field = c("provider", "model", "model_id", "base_url"),
  value = c(
    if (is.null(backend$provider)) {
      NA_character_
    } else {
      as.character(backend$provider)
    },
    if (is.null(backend$model)) NA_character_ else as.character(backend$model),
    if (is.null(info$model_id)) NA_character_ else as.character(info$model_id),
    if (is.null(info$base_url)) NA_character_ else as.character(info$base_url)
  ),
  stringsAsFactors = FALSE
)
```

### Model references

The model should be interpretable and traceable. Use these documentation pages
to review model behavior and limits:

```{r model-reference-links}
provider <- if (is.null(backend$provider)) {
  ""
} else {
  tolower(as.character(backend$provider))
}
model_name <- if (is.null(backend$model)) {
  NA_character_
} else {
  as.character(backend$model)
}
hf_url <- if (!is.na(model_name) && nzchar(model_name)) {
  paste0("https://huggingface.co/", model_name)
} else {
  NA_character_
}

docs <- data.frame(
  reference = c(
    "Hugging Face Inference API",
    "Selected HF model card",
    "OpenAI Embeddings Guide",
    "OpenAI model reference"
  ),
  url = c(
    "https://huggingface.co/docs/api-inference/index",
    hf_url,
    "https://platform.openai.com/docs/guides/embeddings",
    "https://platform.openai.com/docs/models"
  ),
  when_relevant = c(
    "provider = hf",
    "provider = hf",
    "provider = openai",
    "provider = openai"
  ),
  stringsAsFactors = FALSE
)

docs
```

### Where this is stored

- Input configuration file: `demo_backend.yaml`
- Runtime model metadata (written after embedding):
  `project/embeddings/model_id=<...>/embed_model.yaml`

### How to change the model

#### Option A: edit YAML directly

Edit `demo_backend.yaml` before rendering. Typical fields are provider, model,
and endpoint/base URL.

#### Option B: regenerate YAML from R

Hugging Face example:

```{r model-config-example-hf}
backend_hf <- embedding_backend_config(
  provider = "hf",
  model = "BAAI/bge-small-en-v1.5"
)
embedding_backend_save(backend_hf, fn = "demo_backend.yaml")
```

OpenAI example (token must be available in `OVC_API_TOKEN`):

```{r model-config-example-openai}
#| eval: false

Sys.setenv(OVC_API_TOKEN = "<your-openai-token>")

backend_openai <- embedding_backend_config(
  provider = "openai",
  model = "text-embedding-3-small"
)
embedding_backend_save(backend_openai, fn = "demo_backend.yaml")
```

::: {.callout-tip}
Why this matters: embeddings define the geometric space for all downstream
distance and score computations. Changing model changes comparability.
:::

# 2. Dry Run: Text Preparation

This optional dry run executes text preparation and hashing without requesting
embeddings. It writes preview Parquet files so you can inspect what would be
submitted.

::: {.callout-tip}
Use this step to validate custom cleaning logic before spending API calls.
:::

```{r dry-run-embedding-preparation}
embed_corpus(
  project_dir = project_dir,
  backend = backend,
  label = "corpus",
  dry_run = TRUE,
  delete_existing = FALSE,
  verbose = TRUE
)

embed_corpus(
  project_dir = project_dir,
  backend = backend,
  corpus_name = "reference_corpus",
  label = "reference",
  dry_run = TRUE,
  delete_existing = FALSE,
  verbose = TRUE
)

arrow::read_parquet(
  file.path(
    project_dir,
    "reference_corpus_dryrun.parquet"
  )
)

arrow::read_parquet(
  file.path(
    project_dir,
    "corpus_dryrun.parquet"
  )
)
```

# 3. Embed Corpus Label

This step embeds the main corpus and stores vectors under label `corpus`.
`model_id_dir` is captured so downstream distance functions read exactly the
same embedding model partition.

::: {.callout-note}
Embeddings are written to `project/embeddings/model_id=<...>/label=corpus/`.
:::

```{r embed-corpus-label}
model_dir <- embed_corpus(
  project_dir = project_dir,
  backend = backend,
  label = "corpus",
  delete_existing = FALSE,
  verbose = TRUE
)
model_id_dir <- basename(model_dir)
```

# 4. Embed Reference Label

This step embeds `project/reference_corpus` by setting
`corpus_name = "reference_corpus"` and writes results under label `reference`.

::: {.callout-note}
Using labels keeps corpus and reference vectors separate under the same model.
:::

```{r embed-reference-label}
embed_corpus(
  project_dir = project_dir,
  backend = backend,
  corpus_name = "reference_corpus",
  label = "reference",
  delete_existing = FALSE,
  verbose = TRUE
)

```

# 5. Reference-Cosine Distances

`distance_reference_cosine()` computes cosine distances between all corpus and
reference vectors, and adds centroid axes. This creates a full distance matrix
for transparent inspection.

::: {.callout-note}
Distance values are lower for closer vectors.
:::

```{r compute-reference-cosine-distance}
distance_dir <- distance_reference_cosine(
  project_dir = project_dir,
  embeddings_dir = model_id_dir,
  corpus_label = "corpus",
  reference_label = "reference",
  verbose = TRUE
)

arrow::read_parquet(
  file.path(distance_dir, "pairwise-cosine.parquet")
)
```

# 6. Reference-Cosine Scores

Scoring converts distances into a more interpretable ranking signal.
With `method = "linear"`, score = `1 - distance` for every matrix cell.

::: {.callout-note}
Scores are a post-processing step; raw distances stay unchanged on disk.
:::

```{r score-reference-cosine-distance}
score_reference_dir <- score_reference_cosine(
  distance_parquet = distance_dir,
  method = "linear",
  verbose = TRUE
)

arrow::open_dataset(
  score_reference_dir,
  factory_options = list(exclude_invalid_files = TRUE)
) |>
  dplyr::collect()
```

# 7. Ridge Area Distances and Scores

`distance_ridge()` models the reference label as an embedding area
(center + covariance) and calculates distance of each corpus item to that area.
`score_ridge()` then maps area distance to `relevance_score`.

::: {.callout-note}
This gives a second relevance view based on reference-area geometry.
:::

```{r compute-and-score-ridge-distance}
distance_ridge_dir <- distance_ridge(
  project_dir = project_dir,
  reference_label = "reference",
  corpus_label = "corpus",
  verbose = TRUE
)

score_ridge_dir <- score_ridge(
  distance_parquet = distance_ridge_dir,
  verbose = TRUE
)

arrow::open_dataset(
  score_ridge_dir,
  factory_options = list(exclude_invalid_files = TRUE)
) |>
  dplyr::collect()
```

# 8. Output Structure

The final listing helps verify the output layout and where each artifact was
written (`embeddings`, `distance_reference_cosine`, `score_reference_cosine`,
`distance_ridge`, `score_ridge`).

```{r list-demo-output-files}
print_tree <- function(path, prefix = "") {
  entries <- list.files(path, all.files = FALSE, full.names = TRUE, no.. = TRUE)
  entries <- entries[order(tolower(basename(entries)))]
  if (!length(entries)) {
    return(invisible(NULL))
  }

  for (i in seq_along(entries)) {
    entry <- entries[[i]]
    is_last <- i == length(entries)
    branch <- if (is_last) "└── " else "├── "
    cat(
      prefix,
      branch,
      basename(entry),
      if (dir.exists(entry)) "/" else "",
      "\n",
      sep = ""
    )

    if (dir.exists(entry)) {
      next_prefix <- paste0(prefix, if (is_last) "    " else "│   ")
      print_tree(entry, next_prefix)
    }
  }
  invisible(NULL)
}

cat(basename(project_dir), "/\n", sep = "")
print_tree(project_dir)
```
