---
title: "openalexVectorComp Demo Analysis"
subtitle: "Reference vs Corpus Tutorial Workflow"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    df-print: paged
    code-fold: show
    code-summary: "Show code"
    anchor-sections: true
    embed-resources: true
    theme: litera
    include-in-header:
      text: |
        <style>
        main.content {
          max-width: 1000px;
        }
        .quarto-title-block .quarto-title-banner {
          background: linear-gradient(120deg, #143451 0%, #20689f 100%);
          color: #ffffff;
        }
        .callout {
          border-radius: 10px;
          border-width: 1px;
        }
        .cell-output pre, pre {
          white-space: pre-wrap;
          overflow-wrap: anywhere;
          word-break: break-word;
        }
        .cell-output pre {
          background: #f5f8fd;
          border: 1px solid #dbe7f7;
          border-radius: 8px;
          padding: 0.75rem;
        }
        table td, table th {
          vertical-align: top;
        }
        </style>
execute:
  echo: true
  warning: true
  message: true
---

This tutorial walks through a complete `openalexVectorComp` pipeline from raw
corpus/reference text to distance and score outputs. It is intentionally
step-by-step so you can both understand what each function does and reuse the
pattern in your own project.

```{r demo-version-info, echo=FALSE}
pkg_version <- tryCatch(
  as.character(utils::packageVersion("openalexVectorComp")),
  error = function(e) "development"
)
cat("**openalexVectorComp version:** ", pkg_version, "\n\n", sep = "")
```

## Who this tutorial is for

Use this tutorial if you want to:

- understand the practical role of each pipeline function,
- compare direct embedding with optional OpenAI async batch embedding,
- adapt the project structure and workflow to your own datasets.

## Mental model of the pipeline

The workflow has four layers:

1. prepare and validate text,
2. embed corpus/reference text,
3. compute vector distances,
4. transform distances into scores for ranking/filtering.

# Setup

We start by loading the package, reading backend config, and validating input
directories. This makes early failures explicit and avoids expensive requests
when setup is incomplete.

```{r setup-environment}
library(openalexVectorComp)

backend <- embedding_backend_read(file.path(".", "demo_backend.yaml"))

project_dir <- file.path(".", "project")
corpus_dir <- file.path(project_dir, "corpus")
reference_dir <- file.path(project_dir, "reference_corpus")

stopifnot(dir.exists(corpus_dir))
stopifnot(dir.exists(reference_dir))
```

## Active model/backend configuration

This table is your first checkpoint when results look unexpected.

```{r model-configuration-info}
info <- embedding_backend_info(backend)

data.frame(
  field = c("provider", "model", "model_id", "base_url"),
  value = c(
    if (is.null(backend$provider)) NA_character_ else as.character(backend$provider),
    if (is.null(backend$model)) NA_character_ else as.character(backend$model),
    if (is.null(info$model_id)) NA_character_ else as.character(info$model_id),
    if (is.null(info$base_url)) NA_character_ else as.character(info$base_url)
  ),
  stringsAsFactors = FALSE
)
```

## Model references and documentation links

```{r model-reference-links}
provider <- if (is.null(backend$provider)) {
  ""
} else {
  tolower(as.character(backend$provider))
}
model_name <- if (is.null(backend$model)) {
  NA_character_
} else {
  as.character(backend$model)
}
hf_url <- if (!is.na(model_name) && nzchar(model_name)) {
  paste0("https://huggingface.co/", model_name)
} else {
  NA_character_
}

docs <- data.frame(
  reference = c(
    "Hugging Face Inference API",
    "Selected HF model card",
    "OpenAI Embeddings Guide",
    "OpenAI model reference"
  ),
  url = c(
    "https://huggingface.co/docs/api-inference/index",
    hf_url,
    "https://platform.openai.com/docs/guides/embeddings",
    "https://platform.openai.com/docs/models"
  ),
  when_relevant = c(
    "provider = hf",
    "provider = hf",
    "provider = openai",
    "provider = openai"
  ),
  stringsAsFactors = FALSE
)

docs
```

## How model choice affects cost and runtime

Main runtime/cost drivers:

- row count,
- text length per row (tokens),
- re-embedding frequency.

Practical strategy:

- run dry-run first to validate cleaning and avoid unnecessary calls,
- keep labels separate (`corpus`, `reference`, optional `corpus_batch`) so
  experiments do not overwrite each other.

# Dry run for text preparation

`embed_corpus(..., dry_run = TRUE)` runs preprocessing and hashing without
requesting embeddings. Use this to validate text quality before spending API
costs.

```{r dry-run-embedding-preparation}
embed_corpus(
  project_dir = project_dir,
  backend = backend,
  label = "corpus",
  dry_run = TRUE,
  delete_existing = FALSE,
  verbose = TRUE
)

embed_corpus(
  project_dir = project_dir,
  backend = backend,
  corpus_name = "reference_corpus",
  label = "reference",
  dry_run = TRUE,
  delete_existing = FALSE,
  verbose = TRUE
)
```

```{r inspect-dry-run}
arrow::read_parquet(file.path(project_dir, "reference_corpus_dryrun.parquet"))
arrow::read_parquet(file.path(project_dir, "corpus_dryrun.parquet"))
```

# Direct embedding: corpus

`embed_corpus()` reads input parquet, calls the configured backend, and writes
canonical embedding parquet shards under `project/embeddings/...`.

```{r embed-corpus-label}
model_dir <- embed_corpus(
  project_dir = project_dir,
  backend = backend,
  label = "corpus",
  delete_existing = FALSE,
  verbose = TRUE
)
model_id_dir <- basename(model_dir)
```

# Direct embedding: reference

We embed the reference set separately using `label = "reference"`.

```{r embed-reference-label}
embed_corpus(
  project_dir = project_dir,
  backend = backend,
  corpus_name = "reference_corpus",
  label = "reference",
  delete_existing = FALSE,
  verbose = TRUE
)
```

# Optional OpenAI batch embedding and direct-vs-batch comparison

If provider is OpenAI, this section submits async batch embedding for the same
corpus under `label = "corpus_batch"` and attempts a short status/collect loop.
If jobs are still pending, render continues and prints the exact finalize
command for later.

The batch helpers have different roles:

- `embed_corpus_submit_openai_batch()`: enqueue jobs and return quickly.
- `embed_corpus_status_openai_batch()`: refresh/report remote job states.
- `embed_corpus_collect_openai_batch()`: download completed outputs and write
  local batch embedding parquet files.

`finalize_demo_openai_batch()` combines status + collect + compare and returns
`comparison_ready = FALSE` instead of error when not ready yet.

::: {.callout-tip}
After render, check and finalize OpenAI batch jobs with:
```r
openalexVectorComp::embed_corpus_status_openai_batch(
  project_dir = "./project",
  label = "corpus_batch",
  refresh_remote = TRUE
)

openalexVectorComp::finalize_demo_openai_batch(
  demo_dir = ".",
  api_key = keyring::key_get("API_openai_ipbes"),
  label = "corpus_batch"
)
```
:::

```{r openai-batch-compare}
provider_name <- tolower(as.character(if (is.null(backend$provider)) "" else backend$provider))

if (!identical(provider_name, "openai")) {
  message("Skipping OpenAI batch section because provider is: ", provider_name)
} else {
  batch_label <- "corpus_batch"

  submit_info <- embed_corpus_submit_openai_batch(
    project_dir = project_dir,
    backend = backend,
    corpus_name = "corpus",
    label = batch_label,
    delete_existing = TRUE,
    verbose = TRUE
  )
  submit_info

  max_polls <- 20L
  poll_sleep_sec <- 3
  status_df <- data.frame()
  collect_info <- NULL

  for (i in seq_len(max_polls)) {
    status_df <- embed_corpus_status_openai_batch(
      project_dir = project_dir,
      label = batch_label,
      refresh_remote = TRUE
    )

    collect_info <- embed_corpus_collect_openai_batch(
      project_dir = project_dir,
      backend = backend,
      label = batch_label,
      verbose = TRUE
    )

    if (isTRUE(collect_info$downloaded_jobs > 0L)) break

    statuses <- tolower(as.character(status_df$status))
    if (length(statuses) && all(statuses %in% c("completed", "failed", "expired", "cancelled"))) {
      break
    }

    Sys.sleep(poll_sleep_sec)
  }

  status_df
  collect_info

  finalize_out <- finalize_demo_openai_batch(
    demo_dir = ".",
    label = batch_label,
    refresh_remote = TRUE,
    verbose = TRUE
  )

  if (!isTRUE(finalize_out$comparison_ready)) {
    message(finalize_out$message)
    message(
      "Run later: openalexVectorComp::finalize_demo_openai_batch(\"",
      normalizePath(getwd(), mustWork = FALSE),
      "\", api_key = keyring::key_get(\"API_openai_ipbes\"))"
    )
  } else {
    finalize_out$summary_df
    head(finalize_out$comparison_df, 10)
  }
}
```

# Reference-cosine distance

`distance_reference_cosine()` computes a full distance matrix between corpus
and reference vectors, including centroid axes.

```{r compute-reference-cosine-distance}
distance_dir <- distance_reference_cosine(
  project_dir = project_dir,
  embeddings_dir = model_id_dir,
  corpus_label = "corpus",
  reference_label = "reference",
  verbose = TRUE
)

arrow::read_parquet(file.path(distance_dir, "pairwise-cosine.parquet"))
```

# Reference-cosine scores

`score_reference_cosine()` converts distances to ranking-friendly scores.

```{r score-reference-cosine-distance}
score_reference_dir <- score_reference_cosine(
  distance_parquet = distance_dir,
  method = "linear",
  verbose = TRUE
)

arrow::open_dataset(
  score_reference_dir,
  factory_options = list(exclude_invalid_files = TRUE)
) |>
  dplyr::collect()
```

# Ridge area distance and score

`distance_ridge()` models the reference set as an embedding area and computes
how far each corpus row lies from that area. `score_ridge()` maps that distance
to `relevance_score`.

```{r compute-and-score-ridge-distance}
distance_ridge_dir <- distance_ridge(
  project_dir = project_dir,
  reference_label = "reference",
  corpus_label = "corpus",
  verbose = TRUE
)

score_ridge_dir <- score_ridge(
  distance_parquet = distance_ridge_dir,
  verbose = TRUE
)

arrow::open_dataset(
  score_ridge_dir,
  factory_options = list(exclude_invalid_files = TRUE)
) |>
  dplyr::collect()
```

# Output structure and how to reuse it

This tree makes the artifact structure explicit. Reuse the same structure in
your own projects so downstream steps can be automated reliably.

```{r list-demo-output-files}
print_tree <- function(path, prefix = "") {
  entries <- list.files(path, all.files = FALSE, full.names = TRUE, no.. = TRUE)
  entries <- entries[order(tolower(basename(entries)))]
  if (!length(entries)) {
    return(invisible(NULL))
  }

  for (i in seq_along(entries)) {
    entry <- entries[[i]]
    is_last <- i == length(entries)
    branch <- if (is_last) "└── " else "├── "
    cat(
      prefix,
      branch,
      basename(entry),
      if (dir.exists(entry)) "/" else "",
      "\n",
      sep = ""
    )

    if (dir.exists(entry)) {
      next_prefix <- paste0(prefix, if (is_last) "    " else "│   ")
      print_tree(entry, next_prefix)
    }
  }
  invisible(NULL)
}

cat(basename(project_dir), "/\n", sep = "")
print_tree(project_dir)
```
