---
title: "openalexVectorComp OpenAI Demo"
subtitle: "Direct vs Batch Embedding Workflow"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    df-print: paged
    code-fold: show
    code-summary: "Show code"
    anchor-sections: true
    embed-resources: true
    theme: litera
    include-in-header:
      text: |
        <style>
        main.content {
          max-width: 1000px;
        }
        .quarto-title-block .quarto-title-banner {
          background: linear-gradient(120deg, #0f2f4f 0%, #1a5b8f 100%);
          color: #ffffff;
        }
        .callout {
          border-radius: 10px;
          border-width: 1px;
        }
        .cell-output pre, pre {
          white-space: pre-wrap;
          overflow-wrap: anywhere;
          word-break: break-word;
        }
        .cell-output pre {
          background: #f5f8fd;
          border: 1px solid #dbe7f7;
          border-radius: 8px;
          padding: 0.75rem;
        }
        table td, table th {
          vertical-align: top;
        }
        </style>
execute:
  echo: true
  warning: true
  message: true
---

This OpenAI demo shows both embedding paths:

1. direct synchronous embedding (immediate vectors),
2. asynchronous OpenAI batch embedding (submit now, finalize later).

The goal is to make the trade-off explicit: direct is immediate and simple;
batch is built for large workloads and may finish later.

## Who this tutorial is for

Use this tutorial if you want to:

- understand what each pipeline function does (not just run it),
- decide when to use direct embedding vs async batch,
- transfer this workflow to your own corpus/reference project layout.

## Prerequisites

You need:

- a valid OpenAI API key,
- `demo_backend.yaml` configured for provider `openai`,
- corpus files in `project/corpus` and `project/reference_corpus`.

## Mental model of the workflow

The full flow has four layers:

1. **Prepare text** (`dry_run`) to verify input quality and avoid paying for bad text.
2. **Embed text** either directly (immediate) or via async batch (deferred).
3. **Compute distances** between corpus and reference vectors.
4. **Score and inspect** outputs for ranking/relevance workflows.

```{r demo-version-info, echo=FALSE}
pkg_version <- tryCatch(
  as.character(utils::packageVersion("openalexVectorComp")),
  error = function(e) "development"
)
cat("**openalexVectorComp version:** ", pkg_version, "\n\n", sep = "")
```

# Setup

We load package config from `demo_backend.yaml`, assert this is an OpenAI demo,
and verify required input folders exist.

This keeps failures early and explicit. If setup fails, nothing expensive has
been called yet.

```{r setup-environment}
library(openalexVectorComp)

backend <- backend_read(file.path(".", "demo_backend.yaml"))
provider <- tolower(as.character(if (is.null(backend$provider)) "" else backend$provider))
stopifnot(identical(provider, "openai"))

project_dir <- file.path(".", "project")
corpus_dir <- file.path(project_dir, "corpus")
reference_dir <- file.path(project_dir, "reference_corpus")

stopifnot(dir.exists(corpus_dir))
stopifnot(dir.exists(reference_dir))
```

## Active OpenAI configuration

This is the backend used for all calls in this report.
In your own projects, this table is the first thing to check when outputs look
unexpected.

```{r model-configuration-info}
info <- backend_info(backend)

data.frame(
  field = c("provider", "model", "model_id", "base_url"),
  value = c(
    as.character(backend$provider),
    as.character(backend$model),
    as.character(info$model_id),
    if (is.null(info$base_url)) NA_character_ else as.character(info$base_url)
  ),
  stringsAsFactors = FALSE
)
```

## Direct vs Batch at a glance

::: {.callout-important}
Use direct embedding for fast iteration and small jobs.  
Use batch for larger production loads where delayed completion is acceptable.
:::

| Dimension | Direct embedding | Batch embedding |
|---|---|---|
| Latency | Immediate response per call | Delayed completion (async) |
| Operational pattern | Single request/response loop | Submit, poll status, collect later |
| Failure mode | Immediate HTTP failure | Job lifecycle states (`in_progress`, `completed`, etc.) |
| Cost profile | Standard online request path | Often better suited for bulk cost/throughput trade-offs |
| Best use | Development, quick checks, smaller corpora | Larger corpora, scheduled/offline processing |

## Cost and operational considerations

::: {.callout-note}
Exact pricing changes over time. Always verify on the official OpenAI pricing
page before production runs.
:::

Main cost drivers in this workflow:

- number of rows sent for embedding,
- text length per row (token count),
- re-embedding frequency when source text changes.

Practical controls used here:

- run `dry_run` first to validate cleaning and reduce waste,
- separate labels (`corpus`, `reference`, `corpus_batch`) so you can compare
  methods without overwriting,
- use async batch for larger volumes where latency is less important than
  throughput and stability.

# Dry run for text preparation

Before spending API calls, we run the same preprocessing and hashing logic in
`dry_run` mode to inspect what will be embedded.

What to look for in the dry-run tables:

- Is text too short/too noisy for your use case?
- Are there duplicates or obvious cleanup issues?
- Does the text composition (title + abstract) match your retrieval goal?

```{r dry-run-embedding-preparation}
embed_corpus(
  project_dir = project_dir,
  backend = backend,
  label = "corpus",
  dry_run = TRUE,
  delete_existing = FALSE,
  verbose = TRUE
)

embed_corpus(
  project_dir = project_dir,
  backend = backend,
  corpus_name = "reference_corpus",
  label = "reference",
  dry_run = TRUE,
  delete_existing = FALSE,
  verbose = TRUE
)
```

```{r inspect-dry-run}
arrow::read_parquet(file.path(project_dir, "reference_corpus_dryrun.parquet"))
arrow::read_parquet(file.path(project_dir, "corpus_dryrun.parquet"))
```

# Direct embedding (corpus and reference)

We first create direct embeddings. This gives immediate vectors and a stable
baseline for later direct-vs-batch comparison.

`embed_corpus()` does three things:

1. reads the parquet input partition (`corpus` or `reference_corpus`),
2. calls the configured backend to generate vectors,
3. writes canonical parquet shards under
   `project/embeddings/model_id=.../label=.../`.

This direct path is the easiest to debug because failures are immediate.

```{r embed-corpus-label}
model_dir <- embed_corpus(
  project_dir = project_dir,
  backend = backend,
  label = "corpus",
  delete_existing = FALSE,
  verbose = TRUE
)
model_id_dir <- basename(model_dir)
```

```{r embed-reference-label}
embed_corpus(
  project_dir = project_dir,
  backend = backend,
  corpus_name = "reference_corpus",
  label = "reference",
  delete_existing = FALSE,
  verbose = TRUE
)
```

# OpenAI batch workflow and comparison

This section submits a batch job for the same corpus under label
`corpus_batch`. It then performs a short polling window and tries collection.

If OpenAI has not finished yet, the render does not fail. It prints a concrete
follow-up command so you can finalize later.

This is deliberate: batch is asynchronous by design. A robust production flow
must tolerate "not ready yet" states.

## What each batch function does

The async workflow has three separate responsibilities:

- `batch_submit_openai()`: creates batch request files, uploads
  them, and returns quickly with tracked job ids.
- `batch_status_openai()`: refreshes and reports remote job
  states (`in_progress`, `completed`, `failed`, ...), but does not write
  embeddings.
- `batch_collect_openai()`: downloads completed batch outputs and
  writes canonical embedding parquet files under
  `project/embeddings/model_id=.../label=corpus_batch/...`.

`demo_finalize_openai_batch()` wraps the status + collect + comparison steps
for convenience. If jobs are still pending, it returns a non-error
`comparison_ready = FALSE`.

Interpretation guide:

- `status` answers: *What is OpenAI currently doing with my job?*
- `collect` answers: *Did completed jobs produce local embedding files yet?*
- `finalize` answers: *Can I already compare direct and batch vectors?*

::: {.callout-tip}
Use these commands after render to check and finalize batch jobs:
```r
openalexVectorComp::batch_status_openai(
  project_dir = "./project",
  label = "corpus_batch",
  refresh_remote = TRUE
)

openalexVectorComp::demo_finalize_openai_batch(
  demo_dir = ".",
  api_key = keyring::key_get("API_openai_ipbes"),
  label = "corpus_batch"
)
```
:::

```{r openai-batch-compare}
batch_label <- "corpus_batch"

submit_info <- batch_submit_openai(
  project_dir = project_dir,
  backend = backend,
  corpus_name = "corpus",
  label = batch_label,
  delete_existing = TRUE,
  verbose = TRUE
)
submit_info

max_polls <- 20L
poll_sleep_sec <- 3
status_df <- data.frame()
collect_info <- NULL

for (i in seq_len(max_polls)) {
  status_df <- batch_status_openai(
    project_dir = project_dir,
    label = batch_label,
    refresh_remote = TRUE
  )

  collect_info <- batch_collect_openai(
    project_dir = project_dir,
    backend = backend,
    label = batch_label,
    verbose = TRUE
  )

  if (isTRUE(collect_info$downloaded_jobs > 0L)) break

  statuses <- tolower(as.character(status_df$status))
  if (length(statuses) && all(statuses %in% c("completed", "failed", "expired", "cancelled"))) {
    break
  }

  Sys.sleep(poll_sleep_sec)
}

status_df
collect_info

finalize_out <- demo_finalize_openai_batch(
  demo_dir = ".",
  label = batch_label,
  refresh_remote = TRUE,
  verbose = TRUE
)

if (!isTRUE(finalize_out$comparison_ready)) {
  message(finalize_out$message)
  message(
    "Run later: openalexVectorComp::demo_finalize_openai_batch(",
    "\"", normalizePath(getwd(), mustWork = FALSE), "\"",
    ", api_key = keyring::key_get(\"API_openai_ipbes\"))"
  )
} else {
  finalize_out$summary_df
  head(finalize_out$comparison_df, 10)
}
```

After this chunk:

- if `comparison_ready = TRUE`, review summary similarity values and max
  absolute differences to assess numerical parity;
- if `comparison_ready = FALSE`, run the finalize command later and proceed
  with direct embeddings in the meantime.

## Why tiny differences are expected

When comparison is available, you may see values like:

- cosine similarity very close to 1 (for example `0.999999...`),
- max absolute difference very close to 0 (for example `0.000122`).

This is expected. Exact `1` and `0` only occur when vectors are bit-identical.
Small differences usually come from floating-point rounding and different
execution paths in distributed systems.

```{r inspect-direct-vs-batch-differences}
comparison_path <- file.path(
  project_dir,
  "openai_batch_comparison",
  "label=corpus_batch",
  "comparison.parquet"
)

if (!file.exists(comparison_path)) {
  message("No comparison parquet found yet. Run finalize again after batch completion.")
} else {
  cmp <- arrow::read_parquet(comparison_path)

  cat("Distinct cosine similarities (default print):\n")
  print(dplyr::distinct(cmp, cosine_similarity))
  cat("\nDistinct max absolute differences:\n")
  print(dplyr::distinct(cmp, max_abs_diff))

  cat("\nTop rows by max_abs_diff:\n")
  print(
    cmp |>
      dplyr::arrange(dplyr::desc(max_abs_diff)) |>
      dplyr::select(id, cosine_similarity, max_abs_diff) |>
      head(10)
  )

  old_digits <- getOption("digits")
  options(digits = 17)
  on.exit(options(digits = old_digits), add = TRUE)
  cat("\nHigh-precision unique values:\n")
  print(unique(cmp$cosine_similarity))
  print(unique(cmp$max_abs_diff))
}
```

# Distance and score workflow

The remaining steps are unchanged from the core workflow: compute distances and
scores from direct corpus/reference embeddings.

These steps are shown even when batch is pending because your project can
continue with the direct path while batch catches up.

```{r compute-reference-cosine-distance}
distance_dir <- distance_reference_cosine(
  project_dir = project_dir,
  embeddings_dir = model_id_dir,
  corpus_label = "corpus",
  reference_label = "reference",
  verbose = TRUE
)

arrow::read_parquet(file.path(distance_dir, "pairwise-cosine.parquet"))
```

```{r score-reference-cosine-distance}
score_reference_dir <- score_reference_cosine(
  distance_parquet = distance_dir,
  method = "linear",
  verbose = TRUE
)

arrow::open_dataset(
  score_reference_dir,
  factory_options = list(exclude_invalid_files = TRUE)
) |>
  dplyr::collect()
```

```{r compute-and-score-ridge-distance}
distance_ridge_dir <- distance_ridge(
  project_dir = project_dir,
  reference_label = "reference",
  corpus_label = "corpus",
  verbose = TRUE
)

score_ridge_dir <- score_ridge(
  distance_parquet = distance_ridge_dir,
  verbose = TRUE
)

arrow::open_dataset(
  score_ridge_dir,
  factory_options = list(exclude_invalid_files = TRUE)
) |>
  dplyr::collect()
```

# Output structure

This final tree view shows exactly where each artifact was written.

When applying this to your own project, keep this directory structure pattern:

- `embeddings/` for vector artifacts,
- `openai_batch/` for batch job manifests/outputs,
- `openai_batch_comparison/` for direct-vs-batch checks,
- `distance_*` and `score_*` for downstream modeling.

```{r list-demo-output-files}
print_tree <- function(path, prefix = "") {
  entries <- list.files(path, all.files = FALSE, full.names = TRUE, no.. = TRUE)
  entries <- entries[order(tolower(basename(entries)))]
  if (!length(entries)) return(invisible(NULL))

  for (i in seq_along(entries)) {
    entry <- entries[[i]]
    is_last <- i == length(entries)
    branch <- if (is_last) "└── " else "├── "
    cat(prefix, branch, basename(entry), if (dir.exists(entry)) "/" else "", "\n", sep = "")

    if (dir.exists(entry)) {
      next_prefix <- paste0(prefix, if (is_last) "    " else "│   ")
      print_tree(entry, next_prefix)
    }
  }
  invisible(NULL)
}

cat(basename(project_dir), "/\n", sep = "")
print_tree(project_dir)
```
