---
title: "openalexVectorComp Demo Analysis"
format:
  html:
    toc: true
    df-print: paged
    theme: flatly
    include-in-header:
      text: |
        <style>
        body {
          background: #f4f7fb;
        }
        .quarto-title-block .quarto-title-banner {
          background: #1f2a44;
          color: #ffffff;
        }
        .cell-output pre, pre, code {
          white-space: pre-wrap;
          overflow-wrap: anywhere;
          word-break: break-word;
        }
        .cell-output pre {
          background: #eef2ff;
          border: 1px solid #d9e0ff;
          border-radius: 6px;
          padding: 0.75rem;
        }
        </style>
execute:
  echo: true
  warning: true
  message: true
---

This demo runs a complete reference-vs-corpus workflow from the current
project directory. It is designed to be inspectable, so each step writes files
to disk and keeps them for review.

## Setup

In setup we load the package, read backend configuration, and verify that both
`corpus/` and `reference_corpus/` are present. This ensures later steps fail
early with clear causes if inputs are missing.

```{r setup-environment}
library(openalexVectorComp)

demo_dir <- getwd()
backend <- embedding_backend_read(file.path(demo_dir, "demo_backend.yaml"))
project_dir <- file.path(demo_dir, "project")

corpus_dir <- file.path(project_dir, "corpus")
reference_dir <- file.path(project_dir, "reference_corpus")

stopifnot(dir.exists(corpus_dir))
stopifnot(dir.exists(reference_dir))

embed_reference_with_swap <- function(project_dir, backend, dry_run = FALSE) {
  corpus_dir <- file.path(project_dir, "corpus")
  reference_dir <- file.path(project_dir, "reference_corpus")
  corpus_backup <- file.path(project_dir, "_demo_corpus_backup")

  if (dir.exists(corpus_backup)) {
    unlink(corpus_backup, recursive = TRUE)
  }

  ok <- file.rename(corpus_dir, corpus_backup)
  if (!isTRUE(ok)) {
    stop("Could not move corpus directory to temporary backup.")
  }
  on.exit({
    if (dir.exists(corpus_dir)) unlink(corpus_dir, recursive = TRUE)
    if (dir.exists(corpus_backup)) file.rename(corpus_backup, corpus_dir)
  }, add = TRUE)

  dir.create(corpus_dir, recursive = TRUE, showWarnings = FALSE)
  ref_entries <- list.files(reference_dir, full.names = TRUE, no.. = TRUE)
  ok <- if (length(ref_entries)) {
    file.copy(ref_entries, corpus_dir, recursive = TRUE)
  } else {
    logical(0)
  }
  if (length(ref_entries) && !all(ok)) {
    stop("Could not copy reference_corpus into temporary corpus directory.")
  }

  embed_corpus(
    project_dir = project_dir,
    backend = backend,
    label = "reference",
    dry_run = dry_run,
    delete_existing = FALSE,
    verbose = TRUE
  )
}
```

## Dry Run: Text Preparation

This optional dry run executes text preparation and hashing without requesting
embeddings. It writes preview Parquet files so you can inspect what would be
submitted.

```{r dry-run-embedding-preparation}
embed_corpus(
  project_dir = project_dir,
  backend = backend,
  label = "corpus",
  dry_run = TRUE,
  delete_existing = FALSE,
  verbose = TRUE
)

embed_reference_with_swap(
  project_dir = project_dir,
  backend = backend,
  dry_run = TRUE
)

info <- embedding_backend_info(backend)
model_part <- gsub("/", "_", info$model_id, fixed = TRUE)
dry_run_corpus <- file.path(
  project_dir,
  paste0("dry_run_cleaning_model_id=", model_part, "_label=corpus.parquet")
)
dry_run_reference <- file.path(
  project_dir,
  paste0("dry_run_cleaning_model_id=", model_part, "_label=reference.parquet")
)

dry_run_corpus_df <- arrow::read_parquet(dry_run_corpus)
dry_run_reference_df <- arrow::read_parquet(dry_run_reference)

dry_run_corpus_df
dry_run_reference_df
```

## Embed Corpus Label

This step embeds the main corpus and stores vectors under label `corpus`.
`model_id_dir` is captured so downstream distance functions read exactly the
same embedding model partition.

```{r embed-corpus-label}
model_dir <- embed_corpus(
  project_dir = project_dir,
  backend = backend,
  label = "corpus",
  delete_existing = FALSE,
  verbose = TRUE
)
model_id_dir <- basename(model_dir)
```

## Embed Reference Label

This step embeds the reference papers under label `reference`. Because
`embed_corpus()` always reads from `project_dir/corpus`, we temporarily replace
`corpus/` with `reference_corpus/` for this step, then restore it.

```{r embed-reference-label}
embed_reference_with_swap(project_dir = project_dir, backend = backend)

```

## Reference-Cosine Distances

`distance_reference_cosine()` computes cosine distances between all corpus and
reference vectors, and adds centroid axes. This creates a full distance matrix
for transparent inspection.

```{r compute-reference-cosine-distance}
distance_dir <- distance_reference_cosine(
  project_dir = project_dir,
  embeddings_dir = model_id_dir,
  corpus_label = "corpus",
  reference_label = "reference",
  verbose = TRUE
)

distance_file <- file.path(distance_dir, "pairwise-cosine.parquet")
distance_matrix <- arrow::read_parquet(distance_file)
distance_matrix_preview <- distance_matrix[, 1:min(12, ncol(distance_matrix)), drop = FALSE]
distance_matrix_preview
```

## Reference-Cosine Scores

Scoring converts distances into a more interpretable ranking signal.
With `method = "linear"`, score = `1 - distance` for every matrix cell.

```{r score-reference-cosine-distance}
score_reference_dir <- score_reference_cosine(
  distance_parquet = distance_file,
  method = "linear",
  verbose = TRUE
)

score_reference_ds <- arrow::open_dataset(
  score_reference_dir,
  factory_options = list(exclude_invalid_files = TRUE)
)
score_reference_preview <- score_reference_ds |>
  dplyr::collect()
score_reference_preview <- score_reference_preview[, 1:min(12, ncol(score_reference_preview)), drop = FALSE]
score_reference_preview
```

## Ridge Area Distances and Scores

`distance_ridge()` models the reference label as an embedding area
(center + covariance) and calculates distance of each corpus item to that area.
`score_ridge()` then maps area distance to `relevance_score`.

```{r compute-and-score-ridge-distance}
distance_ridge_dir <- distance_ridge(
  project_dir = project_dir,
  reference_label = "reference",
  corpus_label = "corpus",
  verbose = TRUE
)

score_ridge_dir <- score_ridge(
  distance_parquet = distance_ridge_dir,
  verbose = TRUE
)

ridge_scores <- arrow::open_dataset(
  score_ridge_dir,
  factory_options = list(exclude_invalid_files = TRUE)
) |>
  dplyr::collect()

ridge_scores
```

## Output Structure

The final listing helps verify the output layout and where each artifact was
written (`embeddings`, `distance_reference_cosine`, `score_reference_cosine`,
`distance_ridge`, `score_ridge`).

```{r list-demo-output-files}
print_tree <- function(path, prefix = "") {
  entries <- list.files(path, all.files = FALSE, full.names = TRUE, no.. = TRUE)
  entries <- entries[order(tolower(basename(entries)))]
  if (!length(entries)) {
    return(invisible(NULL))
  }

  for (i in seq_along(entries)) {
    entry <- entries[[i]]
    is_last <- i == length(entries)
    branch <- if (is_last) "└── " else "├── "
    cat(
      prefix,
      branch,
      basename(entry),
      if (dir.exists(entry)) "/" else "",
      "\n",
      sep = ""
    )

    if (dir.exists(entry)) {
      next_prefix <- paste0(prefix, if (is_last) "    " else "│   ")
      print_tree(entry, next_prefix)
    }
  }
  invisible(NULL)
}

cat(basename(project_dir), "/\n", sep = "")
print_tree(project_dir)
```
