---
title: "backend-architecture"
vignette: >
  %\VignetteIndexEntry{backend-architecture}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    mermaid:
      theme: forest
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    eval: false
---

# Purpose

This vignette documents the embedding backend architecture in `openalexVectorComp`:

- how the dispatch layer works,
- call order during embedding,
- provider-specific responsibilities,
- how to add a new backend safely.

The focus is implementation details, not end-user quickstart usage.

# File Structure

Backend logic is split into one core file and provider-specific files:

- `R/embed_backend_core.R`:
  - `embedding_backend_config()`
  - `embedding_backend_info()`
  - `embedding_backend_embed_texts()`
  - shared helpers (`.embedding_with_retry()`, `.embedding_request_base()`, ...)
- `R/embed_backend_hf.R`:
  - `.embedding_info_hf()`
  - `.embedding_embed_texts_hf()`
- `R/embed_backend_openai.R`:
  - `.embedding_info_openai()`
  - `.embedding_embed_texts_openai()`
- `R/embed_backend_tei.R`:
  - `.embedding_info_tei()`
  - `.embedding_embed_texts_tei()`

# High-Level Backend Flow

```{mermaid}

flowchart TD
  A[embedding_backend_config] --> B[embedding_backend_info]
  A --> C[embedding_backend_embed_texts]
  C --> D{provider}
  D -->|hf| E[.embedding_embed_texts_hf]
  D -->|openai| F[.embedding_embed_texts_openai]
  D -->|tei| G[.embedding_embed_texts_tei]
  E --> H[Matrix V1..Vd]
  F --> H
  G --> H
```

# End-to-End Call Order in `embed_corpus()`

The `embed_corpus()` function is now a pipeline orchestrator that uses the backend layer.

```{mermaid}

sequenceDiagram
  participant User
  participant embed_corpus
  participant Core as embed_backend_core
  participant Provider as provider_adapter
  participant API as embedding_endpoint
  participant FS as parquet_output

  User->>embed_corpus: embed_corpus(project_dir, backend=...)
  embed_corpus->>Core: embedding_backend_config(...) (if backend is NULL)
  embed_corpus->>Core: embedding_backend_info(backend)
  embed_corpus->>FS: Load existing hashes (id + text_hash)
  loop Arrow scan batches
    embed_corpus->>embed_corpus: Build canonical text
    embed_corpus->>embed_corpus: Filter unchanged rows
    embed_corpus->>Core: embedding_backend_embed_texts(texts, backend)
    Core->>Provider: provider dispatch
    Provider->>API: batched HTTP requests
    API-->>Provider: vectors
    Provider-->>Core: matrix
    Core-->>embed_corpus: matrix
    embed_corpus->>FS: write parquet batch
  end
  embed_corpus-->>User: output model directory
```

# Data Contract

## Input dataset expected by `embed_corpus()`

- `id`
- `title`
- `abstract`

## Canonical text construction

For each row:

- if abstract exists:
  - `Title: {title}\nAbstract: {abstract}`
- else:
  - `Title: {title}`

## Output columns per embedded row

- `id`
- `text_hash`
- `provider`
- `model_id`
- `created_at`
- `V1..Vd`

Embeddings are stored by partition path:

- `embeddings/model_id=<...>/label=<...>/batch=<n>/embeddings-*.parquet`

# Shared Backend Rules

All provider adapters should follow these rules:

1. Return a numeric matrix with one row per input text.
2. Set output column names to `V1..Vd`.
3. Raise an error when output row count mismatches input size.
4. Use `.embedding_with_retry()` for transient failures.
5. Use `.embedding_request_base()` so auth is consistent.

# Authentication Model

The backend layer uses one environment variable:

- `OVC_API_TOKEN`

When set, requests include:

- `Authorization: Bearer <OVC_API_TOKEN>`

This keeps auth handling provider-agnostic.

# Provider Responsibilities

## Hugging Face (`provider = "hf"`)

- Default base URL: `https://router.huggingface.co/hf-inference`
- Default model: `BAAI/bge-small-en-v1.5`
- Embedding endpoint shape:
  - `/models/{model}`
- Body shape:
  - `{"inputs": [...]}`.

## OpenAI (`provider = "openai"`)

- Default base URL: `https://api.openai.com/v1`
- Default model: `text-embedding-3-small`
- Embedding endpoint:
  - `/embeddings`
- Body shape:
  - `{"model": "...", "input": [...]}`.

## TEI (`provider = "tei"`)

- Default base URL: `http://localhost:3000`
- Embedding endpoint:
  - `/embed` (or explicit full endpoint via `tei_url`)
- Probes `/info` when available; gracefully degrades when not available.

# Function Examples

## 1) HF backend (default-style)

```{r}
library(openalexVectorComp)

backend <- embedding_backend_config(
  provider = "hf",
  model = "BAAI/bge-small-en-v1.5",
  max_batch_size = 64
)

info <- embedding_backend_info(backend)
emb <- embedding_backend_embed_texts(
  texts = c("Title: A\nAbstract: B", "Title: C\nAbstract: D"),
  backend = backend
)
dim(emb)
```

## 2) OpenAI backend

```{r}
backend <- embedding_backend_config(
  provider = "openai",
  model = "text-embedding-3-small",
  max_batch_size = 256
)
```

## 3) Local TEI backend

```{r}
backend <- embedding_backend_config(
  provider = "tei",
  base_url = "http://localhost:3000",
  max_batch_size = 128
)
```

# How to Add a New Backend

Assume new provider name `"acme"`.

## Step 1: add dispatch entry in core

In `embedding_backend_config()`:

- include `"acme"` in `provider = c(...)`
- define provider defaults in `switch(provider, ...)`

In `embedding_backend_info()`:

- add branch: `acme = .embedding_info_acme(backend)`

In `embedding_backend_embed_texts()`:

- add branch: `acme = .embedding_embed_texts_acme(texts, backend)`

## Step 2: create provider file

Create `R/embed_backend_acme.R`:

- `.embedding_info_acme(backend)`
- `.embedding_embed_texts_acme(texts, backend)`

Use helpers from core:

- `.embedding_with_retry()`
- `.embedding_request_base()`
- `.embedding_as_matrix()`
- `.embedding_batch_starts()`

## Step 3: enforce adapter contract

Before returning, ensure:

1. `nrow(emb) == length(batch)`
2. matrix numeric
3. `colnames(emb) <- paste0("V", seq_len(ncol(emb)))`

## Step 4: docs and validation

- document provider behavior in this vignette and roxygen comments.
- run roxygen:

```{r}
roxygen2::roxygenise(".", load = "source")
```

- parse/check:

```{r}
for (f in list.files("R", pattern = "[.]R$", full.names = TRUE)) {
  parse(f)
}
```

# Operational Flowchart for `embed_corpus()`

```{mermaid}
flowchart TB
  classDef default fill:#f7f9fc,stroke:#4b5563,color:#111827;
  linkStyle default stroke:#374151,stroke-width:2px,color:#374151;
  A[Read corpus batch] --> B[Build canonical text]
  B --> C[Compute text_hash]
  C --> D{hash changed?}
  D -->|No| E[Skip row]
  D -->|Yes| F[Embed via backend]
  F --> G[Validate matrix rows]
  G --> H[Write parquet batch]
  H --> I[Update in-memory hash index]
  E --> J{More batches?}
  I --> J
  J -->|Yes| A
  J -->|No| K[Return model_dir]
```

# Troubleshooting

## 401 / 403 from provider API

- Check `OVC_API_TOKEN`.
- Verify token scope for the selected provider/model.

## Row mismatch errors

- Provider returned fewer/more vectors than inputs.
- Reduce `max_batch_size` and retry.

## Model metadata unavailable (`dim = NA`)

- Some hosted endpoints do not expose model info.
- This is acceptable; downstream code should infer dimensions from embeddings.
