---
title: "simplestart"
vignette: >
  %\VignetteIndexEntry{simplestart}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    eval: false
---


# Introduction

This vignette walks through a minimal end‑to‑end run using openalexVectorComp:

- use a running TEI (Text Embeddings Inference) server endpoint,
- embed a small Parquet corpus into shard files,
- compute prototype margins and ridge distances.

The code mirrors the package functions directly, with more context and guidance.

For detailed text-preparation logic before embedding (including abstract
cleaning rules, policies, and customization), see the vignette
`abstract-cleaning`.

# Requirements

- A running TEI server endpoint, e.g. `http://localhost:3000/embed`.
- Suggested R packages (for vignette rendering): `knitr`, `rmarkdown`, and `quarto`.
- The example data located under `inst/examples/` in this package source.

## Notes on execution

- Chunks in this vignette default to `eval: false` to avoid launching external
  processes during package build. Remove or set `eval: true` to run locally.
- Long‑running steps (embedding) are intentionally small in batch size to keep
  resource usage low when you do run them.

# TEI endpoint

Start TEI outside the package, for example from a shell:

```bash
text-embeddings-router --model-id BAAI/bge-small-en-v1.5 --port 3000
```

For detailed operational guidance (start/stop/health checks), see the vignette
`tei-server-operations`.

# Embed a small corpus

The example corpus ships with the package sources under `inst/examples/corpus/`.
We embed to `inst/examples/embedings/` (note: the folder name mirrors the example script).

```{r}
#| label: embed-corpus
embed_corpus(
  project_dir = "inst/examples",
  backend = embedding_backend_config(
    provider = "tei",
    base_url = "http://localhost:3000"
  ),
  label = "corpus",
  batch_size = 15,
  verbose = TRUE
)

embed_corpus(
  project_dir = "inst/examples",
  backend = embedding_backend_config(
    provider = "tei",
    base_url = "http://localhost:3000"
  ),
  label = "reference",
  batch_size = 15,
  verbose = TRUE
)
```

# Compute prototype margins

Prototype distances are computed pairwise between all vectors in a reference
label partition and all vectors in a corpus label partition.

```{r}
#| label: prototype-margins
distance_prototype(
  project_dir = "inst/examples",
  embeddings_dir = "model_id=BAAI_bge-small-en-v1.5",
  corpus_label = "corpus",
  reference_label = "reference"
)
```

# Compute ridge distances

Ridge distances now model the `reference` label as an embedding area
(centroid + covariance) and score all `corpus` rows by Mahalanobis-style
distance converted to `relevance_score`.

```{r}
#| label: ridge-distances
distance_ridge(
  project_dir = "inst/examples/",
  reference_label = "reference",
  corpus_label = "corpus"
)
```

# Troubleshooting

- Can’t find `text-embeddings-router`:
  - Install the binary and ensure it is on PATH.
- Port in use:
  - Start TEI on another port and update `backend = embedding_backend_config(provider = "tei", base_url = "http://localhost:3001")`.
- Slow embedding or timeouts:
  - Reduce `batch_size`, and verify the server’s `/info` limits.

# Reproducibility

```{r}
#| label: session-info
sessionInfo()
```
