---
title: "simplestart"
vignette: >
  %\VignetteIndexEntry{simplestart}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    eval: false
---


# Introduction

This vignette walks through a minimal end‑to‑end run using openalexAutoTag:

- start a local TEI (Text Embeddings Inference) server,
- embed a small Parquet corpus into shard files,
- compute prototype margins and ridge distances.

The code mirrors `inst/examples/run_autotag_example.R`, with more context and guidance.

# Requirements

- A TEI server binary (`text-embeddings-router`) available on your system PATH,
  or pass the `bin` argument to `tei_start()`.
- Suggested R packages (for vignette rendering): `knitr`, `rmarkdown`, and `quarto`.
- The example data located under `inst/examples/` in this package source.

## Notes on execution

- Chunks in this vignette default to `eval: false` to avoid launching external
  processes during package build. Remove or set `eval: true` to run locally.
- Long‑running steps (embedding) are intentionally small in batch size to keep
  resource usage low when you do run them.

# Start a TEI server

Use `tei_start()` to launch a local TEI server in the background. The process is
tracked by the package and can be stopped with `tei_stop()`. The helper
`tei_state()` shows the tracked state.

```{r}
#| label: start-tei
library(openalexAutoTag)

# Start a TEI server; the URL is remembered and used by defaults elsewhere
tei_start(
  model = "sentence-transformers/all-mpnet-base-v2",
  args = c("--auto-truncate")
)

# Inspect the remembered state (pid/port/url)
tei_state()
```

To get further info, you can use `tei_info()`:

```{r}
#| label: tei-info
tei_info()
```

# Embed a small corpus

The example corpus ships with the package sources under `inst/examples/corpus/`.
We embed to `inst/examples/embedings/` (note: the folder name mirrors the example script).

```{r}
#| label: embed-corpus
tei_embed(
  project_dir = "inst/examples",
  batch_size = 15,
  verbose = TRUE
)
```


# Stop the TEI server

When you are done with the embedding, you can stop the background server. The package also attempts this on
unload via `.onUnload()`.

```{r}
#| label: stop-tei
tei_stop()
```

# Compute prototype margins

Prototype margins require two ID lists: a positive “included” set and a negative
“excluded” set. The example CSVs are in `inst/examples/`.

```{r}
#| label: prototype-margins
prototype_distance(
  project_dir = "inst/examples",
  included = "inst/examples/included_part.csv",
  excluded = "inst/examples/excluded_part.csv",
  batch_size = 10
)
```

# Compute ridge distances

Ridge distances train a regularized logistic model between the two sets and
score the entire dataset. The outputs mirror the embeddings structure.

```{r}
#| label: ridge-distances
ridge_distance(
  project_dir = "inst/examples/",
  included = "inst/examples/included_part.csv",
  excluded = "inst/examples/excluded_part.csv"
)
```

# Troubleshooting

- Can’t find `text-embeddings-router`:
  - Install the binary and ensure it is on PATH, or pass `bin = "/path/to/text-embeddings-router"`.
- Port in use:
  - Use `tei_start(port = 3001)` (and friends). All embedding helpers will use
    the remembered URL by default.
- Slow embedding or timeouts:
  - Reduce `batch_size`, and verify the server’s `/info` limits.

# Reproducibility

```{r}
#| label: session-info
sessionInfo()
```
